---
output:
  word_document: default
  html_document: default
---
# Descriptive Statistics

## Calculating group means

One of the most basic exploratory tasks with any data set involves computing the mean, variance, and other descriptive statistics. A previous section has already demonstrated how to obtain many of these statistics from a data set, using the `summary()`, `mean()`, and `sd()` functions. However these functions were used in the context of an entire data set or column from a data set; in most cases it will be more informative to calculate these statistics for groups of data, such as experimental treatments. The `tapply()` function can be used for this purpose. Summary statistics for the corn irrigation data set were calculated previously using the `summary()` function. The `tapply()` function allows us to calculate similar descriptive statistics for groups of data, most commonly for treatments, but also for other logical groups, such as by experimental sites, or years. The `tapply()` function requires three arguments: 

- The data column you want to summarize (response variable)
- The data column you wish to group the data by (treatment or grouping variable)
- The function you want to calculate (mean, standard deviation, maximum, etc.)

The example below will calculate mean yield for each irrigation block (Full irrigation or Limited irrigation). 
```{r}
corn.dat <- read.csv("http://rstats4ag.org/data/irrigcorn.csv")
tapply(corn.dat$Yield.BuA, corn.dat$Irrig, mean)
```
To specify multiple grouping variables, we use the `list()` function within the `tapply()` function:
```{r}
tapply(corn.dat$Yield.BuA, list(corn.dat$Population.A, corn.dat$Irrig), mean)
```

## Calculating other group statistics with tapply

We can use the same syntax to calculate nearly any statistic for the groups by replacing the `mean` argument with another function, such as `max`, `min`, `sd` etc. 
```{r}
tapply(corn.dat$Yield.BuA, list(corn.dat$Population.A, corn.dat$Irrig), median)
```
You can also specify other functions or operations. For example, there is no default function in `R` to calculate the CV, or coefficient of variation. This statistic is commonly used in crop sciences. The CV is simply the standard deviation divided by the mean, and so we can calculate the CV using by storing the output from `tapply()` into an object that we can call upon later.
```{r}
popirr.sd   <- tapply(corn.dat$Yield.BuA, 
                      list(corn.dat$Population.A, corn.dat$Irrig), sd)
popirr.mean <- tapply(corn.dat$Yield.BuA, 
                      list(corn.dat$Population.A, corn.dat$Irrig), mean)
popirr.cv<-popirr.sd/popirr.mean
round(popirr.cv, 3)
```
The `round()` function simply rounds the result to 3 decimal places to make the output more readable. Alternatively, if the CV is a statistic we will use regularly, it may be worthwhile to write our own function to save keystrokes later on. We can do this using `function()`.
```{r}
cv <- function(x) {
  xm  <- mean(x)
  xsd <- sd(x)
  xcv <- xsd/xm
  round(xcv, 4)
}
```
Or, more efficiently:
```{r}
cv <- function(x) {
  xcv <- round(sd(x) / mean(x), 4)
}
tapply(corn.dat$Yield.BuA, list(corn.dat$Population.A, corn.dat$Irrig), cv)
```
This approach is desirable if the CV will be used repeatedly. Once the function `cv()` has been defined, you can call upon it repeatedly with very few keystrokes.

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 100)
```

## Using dplyr to summarize data

In addition to using `tapply()` from base stats R package, there is an excellent package for summarizing data called `dplyr`. We can get similar information from the corn data set using the `dplyr` package using the following syntax (explained in more detail below).
```{r, message=FALSE }
library(dplyr)
corn.summary <- corn.dat %>%
  group_by(Population.A, Irrig) %>%
  summarize(N        = length(Yield.BuA),
            AvgYield = round(mean(Yield.BuA),1),
            CV       = cv(Yield.BuA))
corn.summary            
```

We will explain and demonstrate some of the basic uses of `dplyr` using herbicide resistant weed data from Canada. The following data set was assembled using data from [weedscience.org](http://www.weedscience.org).

```{r, options(width=200)}
resistance<-read.csv("http://rstats4ag.org/data/CanadaResistance2.csv")
resistance<-na.omit(resistance)
head(resistance)
```

In the data file, there are instances of multiple resistance (resistance to more than one mode of action).
The second line of the code above (`na.omit()`) removes all lines in the data set that contain `NA` (which is the default value in R for missing data). This is necessary for this particular example so that instances of multiple resistance are not counted multiple times. We may want to use the information from those rows later on, however, and therefore this method is preferable to deleting the rows from the raw data file.

One of the first things we might like to do is simply count the number of herbicide resistant weed cases by mode of action. We could do this rather quickly using the `tapply()` function discussed in the previous section: `tapply(resistance$ID, resistance$MOA, length)`. In the code below, we will use the `dplyr` package to get some interesting information from this data set. First, we can simply count the number of herbicide resistant weeds for each mode of action in the data set. Initially it seems like much more typing compared to using `tapply()`, but later on it will become clear why this is a potentially more powerful method.

```{r, message=FALSE, results='hide'}
library(dplyr)
bymoa<-group_by(resistance, MOA)
moatotal<-summarize(bymoa, total=length(ID))
print(moatotal)
```
In the code above, we are loading the dplyr package, then using two functions from that package. The `group_by()` function will provide the information on how the data are grouped. In this case, we want to group the data by MOA. We can then use the the `summarize()` function to calculate means, counts, sums, or any other operation on the data. Above, we used the `length()` function to count the unique instances of the ID variable for each mode of action. Storing each step with the `<-` operator can be a little cumbersome, especially if we will not need to use the intermediary steps again. The `dplyr` package uses the `%>%` operator to string together several steps. The code below will result in the same summary as above, with fewer keystrokes, and without storing the intermediary grouping as a separate object. The first line provides the name of the data set (`resistance`) followed by the `%>%` operator. This tells `R` to expect a function on the next line from the dplyr package. The second line groups the resistance data set by MOA, and the third line summarizes the data by calculating the length of the ID variable for each level of the grouping factor. Effectively, this counts each instance of resistance. The last line prints out the resulting data frame.
```{r}
moatotal <- resistance %>%
  group_by(MOA) %>%
  summarize(total=length(ID))
moatotal
```
The real power and elegance of the `dplyr` package becomes apparent once we begin using multiple groupings, and creating new variables. Below, we generate totals for each mode of action again, but also by year of listing by using the `group_by()` and `sumarize()` functions. 
```{r}
totals <- resistance %>%
  group_by(Year, MOA) %>%
  summarize(total = length(ID)) 
head(totals)
tail(totals)
```
We can then create a new variable using the `mutate()` function; we will add a cumulative total of herbicide resistance cases over time, while retaining information about mode of action by first ungrouping the data, then using mutate to calculate the cumulative total using the `cumsum()` from the default stats package.
```{r}
totals <- resistance %>%
  group_by(Year, MOA) %>%
  summarize(total = length(ID)) %>%
  ungroup() %>%
  mutate(cumulative = cumsum(total))
head(totals)
tail(totals)
```

We can also use `dplyr` to look at other grouping, including by province or weed species, and then use the `subset()` function to print out only one group. In the below example, we will calculate totals for each province by weed species then keep only the cases from Saskatchewan.
```{r}
resistance.SK <- resistance %>%
  group_by(Province, SciName) %>%
  summarize(total=length(ID)) %>%
  filter(Province == "Saskatchewan") %>%
  arrange(-total) 
resistance.SK
```

*****

